Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-15 article.tex
fix typo
article: clean up references to code variable names:  surround with \verb+...+; remove references to no-longer existent Plus type class

\title{Error Reporting Parsers:  a Monad Transformer Approach}
\author{Matt Fenwick\email{mfenwick100@gmail.com}}

@@

literal :: Eq t => t -> Parser t t
literal x = afilter (== x) item
\end{verbatim}
First, we need to define the Parser type, which says that a Parser is a function
that operates on lists of tokens, producing a new list of tokens and a result
value, and possibly failing.
Next is the item parser, the simplest parser we'll deal with.  All it does is 
consume a single token if available, and succeed otherwise.
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports ap and liftM from Control.Monad.
For choosing between alternatives, the MonadPlus type class will do, so we create
an instance that is defined partly in terms of the MonadPlus instance of the
underlying result type, Maybe.
Last are a couple of convenience combinators for creating specific token parsers.

Now for some examples:  first, let's see item in action.  It fails on empty lists,
and succeeds consuming one token otherwise:
\begin{verbatim}
*Main> getParser item ""
Nothing

*Main> getParser item "abcde"
Just ("bcde",'a')
\end{verbatim}
We can run parsers in sequence using the Applicative combinators.  First one parser
is run, then the second.  Note that the token position is handled such that the
second parser sees the token list output by the first parser, and that both parsers
must succeed in order for the entire parser to succeed:
\begin{verbatim}
*Main> getParser (item *> item) "abcde"
Just ("cde",'b')

@@

*Main> getParser (item *> item) "a"
Nothing
\end{verbatim}
Parsers built out of literal only succeed when the next token matches the given
one:
\begin{verbatim}
*Main> getParser (literal 'a') "abcde"
Just ("bcde",'a')

*Main> getParser (literal 'a') "bcde"
Nothing
\end{verbatim}
Using the MonadPlus combinators, we can create parsers that succeed if any of
their sub-parsers succeed.  This is what allows parsers to backtrack, trying
various alternatives if one fails:
\begin{verbatim}
*Main> getParser (literal 'a' <|> literal 'b') "abcde"
Just ("bcde",'a')

@@

First, note that the parser type we used earlier -- [t] -> Maybe ([t], a) -- 
can be built out of the StateT monad transformer applied to Maybe, so the new
parser type is:
\begin{verbatim}
type Parser t a = StateT [t] Maybe a
\end{verbatim}
Now we can build the item parser using the combinators from MonadState, since
its token lists are the 'state' that it modifies.
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}

@@

And a combinator that checks that the next character is equal to a given character:
\begin{verbatim}
literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
literal c = satisfy ((==) c)
\end{verbatim}

@@

munch p = many0 (whitespace <+> comment) *> p
\end{verbatim}
Here are these three parsers in action:
\begin{verbatim}
*Main> runParser whitespace "\n \t   \tabc"
Just ("\n \t   \t","abc")

@@

The comment parser uses a new combinator, not1, which is built out of a new
type class, Switch, which models computations which can be switched from
successful to failing and vice versa:
\begin{verbatim}
class Switch f where
  switch :: f a -> f ()

@@

With munch in hand, we can now fix opencurly and try it out:
\begin{verbatim}
opencurly = munch $ literal '{'

*Main> map (runParser opencurly) [";abc\n  {def", "{def", "(def", "}def"]
[Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
\end{verbatim}
Excellent!  Now the open curly parser correctly throws away whitespace and comments,
and succeeds, consuming one character, when the first non-junk character is an open 
curly brace, and fails otherwise.
It's straightforward to define parsers for the other three brace tokens:
\begin{verbatim}
closecurly = munch $ literal '}'
openparen  = munch $ literal '('
closeparen = munch $ literal ')'
\end{verbatim}

@@

*Main> runParser symbol ";sdfasdfsa\n  123abc"
Nothing
\end{verbatim}
Note the use of many1 to introduce repetition,
and we're again using munch to throw away leading junk.

\subsection{Syntactic structures}
Whereas our token parsers dealt with the syntactic primitives of the Woof grammar, 
the remaining parsers will have to implement grammar rules that combine the tokens
into syntactic structures.
The first combining form is a function application.  The rule says that it's delimited
by matching parentheses, in between which must appear one form as the operator, followed
by any number of forms as the arguments to which the operator is applied.  We can
say that like so:
\begin{verbatim}
application =
    openparen    *>
    pure AApp   <*>
    form        <*>
    many0 form  <*
    closeparen
\end{verbatim}
Note that, for now, we're doing all parser sequencing using the Applicative
combinators -- we don't need to use the monadic ones.  Also, we used pure to
inject a value -- the function AApp -- into the parser.  pure has no effect on
the tokens and always succeeds:
\begin{verbatim}
*Main> runParser (pure 3) ""
Just (3,"")

@@

Next, we move on to define, which is also a straightforward translation from the grammar:
\begin{verbatim}
define =
    opencurly                    *>
    check (== "define") symbol   *>
    pure ADefine                <*>
    symbol                      <*>
    form                        <*
    closecurly
\end{verbatim}

Lambda is more complicated -- not only does it have additional syntax with its parameter list,
but we also need to ensure that the parameter names are unique.  We can do that using the 
check combinator again:
\begin{verbatim}
lambda = 
    opencurly                       *>
    check (== "lambda") symbol      *>
    opencurly                       *>
    pure ALambda                   <*>
    check distinct (many0 symbol)  <*>
    (closecurly                     *>
     many1 form                    <*
     closecurly)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
If the symbols are distinct, the parameter list subparser will succeed, and parsing will
continue; if there's a repeated symbol, the parsing will stop with a failure.

\subsection{Finishing touches}
Now all we have left to do is to define the special, form (which we've already used to build
the previous parsers), and woof parsers.  The special parser is simple:  it parses either 
define or lambda:
\begin{verbatim}
special = define <+> lambda
\end{verbatim}

@@

except that the types don't match -- symbol's type parameter is a String, whereas the other two
have ASTs, so we fix that by mapping the ASymbol function over symbol:
\begin{verbatim}
form = fmap ASymbol symbol <+> application <+> special
\end{verbatim}

The final parser, woof, needs to not only parse all the forms, but also make sure that the
entire input is consumed.  The end of input check is done with the end parser:
\begin{verbatim}
end = switch item
\end{verbatim}

@@

*Main> runParser woof "a b c {define q (f x)},"
Nothing
\end{verbatim}
Note that the first two parses consume the entire input, which the last one 
fails because it can't recognize the trailing comma.

@@

runParser :: Parse e t a -> [t] -> Either e (Maybe (a, [t]))
runParser p xs = runMaybeT (runStateT p xs)
\end{verbatim}
Now we have three different possible results:  success, failure, and an error:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))

@@

*Main> runParser (throwError "oops!") ""
Left "oops!"
\end{verbatim}
Although the choice combinator can recover from failures, it can't recover from
errors.  Compare:
\begin{verbatim}
*Main> runParser (satisfy (== 'a') <|> satisfy (== 'b')) "babc"
Right (Just ('b',"abc"))

@@

\subsection{Type Class Preliminaries}
Unfortunately, the standard MonadError instance for the Either datatype does
not fit our needs -- it requires an Error constraint that we do not want.  To
solve this problem, we'll implement our own MonadError type class, based on the
standard one, and instances of it.  We'll need an instance for each member of
our stack -- Either, StateT, and MaybeT: 
\begin{verbatim}
class Monad m => MonadError e m | m -> e where
  throwError :: e -> m a
  catchError :: m a -> (e -> m a) -> m a

@@

\subsection{Token parsers}
By adding the position to the input, we've created a new problem 
for ourselves: our original token parsers operated on 
character lists, but we now need them to work on token lists.  
The cause of the problem is the \verb+literal+ combinator, so we
replace it with one for token lists:
\begin{verbatim}
character :: (MonadState [Token] m, Plus m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}
And now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')

@@

ocurly = munch $ character '{'
ccurly = munch $ character '}'
oparen = munch $ character '('
cparen = munch $ character ')'
symbol = munch $ many1 char
  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
\end{verbatim}
The only differences were the replacement of literal with character and a couple
uses of the chr accessor.  The munch combinator didn't have to change at all 
(although its inferred type did in fact change).

\subsection{Error messages}
Now we modify the remaining parsers to report accurate and useful errors.
Here's the error messages that we'll use to report each error:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eDefClose   =  "define: missing close curly"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eLamClose   =  "lambda: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

The MonadError type class provides two useful combinators, throwError for 
generating errors, and catchError for dealing with them.  We can ignore 
catchError because we don't need to recover from errors, just report them. 
When should an error be generated and what 
data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
introduce the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Plus m) => e -> m a -> m a
commit err p = p <+> throwError err
\end{verbatim}
This combinator takes two arguments: 1. an error value, and 2. a monadic 
computation.  It first tries to run the computation, and if the computation 
fails, generates an error whose contents are the first argument.  The 
definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+. So 
what data do we want to include in our errors?  A String describing the error 
and a token reporting the position should describe the error decently:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}
Now lets start with \verb+Application+ as our first modified, error-reporting parser:
\begin{verbatim}
application =
    openparen                             >>= \open ->
    commit (eAppOper, open) form          >>= \op ->
    many0 form                            >>= \args ->
    commit (eAppClose, open) closeparen   >>
    return (AApp op args)
\end{verbatim}
This parser works by running the open-parenthesis parser, capturing its value 
in \verb+open+, and continuing with the parsing; if either \verb+form+ or 
close-parenthesis fails, an appropriate error can be generated which includes 
a string and the location of the open parenthesis.  However, if the open 
parenthesis is not successfully parsed, it's a failure but not an error.  
Here are some examples to demonstrate this in action:
\begin{verbatim}
*Main> runParser application (countLineCol "oops")
Right Nothing

@@

*Main> runParser application (countLineCol "(a b")
Left ("application: missing close parenthesis",('(',1,1))
\end{verbatim}
Although the previous version of the \verb+application+ parser was implemented 
with just applicative combinators, this version can't be:  \verb+open+ 
is used to build the following parsers, making this decidedly monadic.

@@

lambda =
    opencurly                                >>= \open ->
    check (== "lambda") symbol               >>
    commit (eLamParam, open) opencurly       >>= \p_open ->
    many0 symbol                             >>= \params ->
    (if distinct params 
        then return ()
        else throwError (eLamDupe, p_open))  >>
    commit (eLamPClose, p_open) closecurly   >>
    commit (eLamBody, open) (many1 form)     >>= \bodies ->
    commit (eLamClose, open) closecurly      >>
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
Note that \verb+commit+ isn't used until we can be sure that we're in the right 
rule:  if, in the \verb+define+ parser, we committed to parsing the "define" 
keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
that failed.  Thus, if treating errors as unrecoverable, it's important not to 
place a \verb+commit+ where backtracking might be necessary to parse a valid 
alternative.  However, this isn't the case with how we've used it, since once 
we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule, and allowing backtracking would only destroy our ability to produce clean 
and accurate error reports.

\subsection{Error messages:  finishing touches}
To complete the error-reporting parser, we need to change a couple more details. 
First, we want an error reported when an open curly brace is found, but no 
special form can be reported.  To accomplish this, we can extend the \verb+special+ 
parser with a third alternative, which results in an error if it can parse
an open curly brace, but in a backtracking failure if not:
\begin{verbatim}
special = define <+> lambda <+> (ocurly >>= \o -> throwError (esName, o))
\end{verbatim}
The \verb+form+ parser doesn't have to be changed:
\begin{verbatim}
form = fmap ASymbol symbol <+> application <+> special
\end{verbatim}
When we can no longer parse any forms, instead of failing, we'd like to report an error if there's any input
left, and otherwise succeed.  To implement this parser, we'll first munch any leading whitespace or comments, 
as usual, and then we'll check the input:  if it's empty, great; if not, report an error at the location of the
first remaining token, along with the appropriate error message that we defined earlier:
\begin{verbatim}
endCheck = 
    many0 (whitespace <+> comment) *>
    get >>= \xs -> case xs of
                        (t:_) -> throwError (ewUnp, t)
                        []    -> pure ()
\end{verbatim}
Now our \verb+woof+ parser uses this \verb+endCheck+ parser after it has consumed all the forms that it can:
\begin{verbatim}
woof = many0 form <* endCheck
\end{verbatim}

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
*Main> runParser woof (countLineCol "{define fgh {lambda {f x} (f x)}}")
Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
                                     [AApp (ASymbol "f") [ASymbol "x"]])],
       []))
\end{verbatim}
Here's the proof showing that the parser correctly report all of the errors 
covered by our spec:
\begin{verbatim}
*Main> let p = runParser woof . countLineCol

@@

runParser' :: Parser' t a -> [t] -> (Maybe a, [t])
runParser' p xs = runIdentity $ runStateT (runMaybeT p) xs
\end{verbatim}
We can compare it to the earlier parser:
\begin{verbatim}
*Main> runParser' (literal 'a') "aqrs"
(Just 'a',"qrs")

@@

*Main> runParser (literal 'a') "aqrs"
Just ('a',"qrs")
\end{verbatim}
The difference is that Maybe/State always returns a modified state, regardless
of the success of the computation, whereas State/Maybe only returns a modified
state if the computation is successful.  This means that backtracking doesn't
work right with Maybe/State -- tokens are consumed even when the parsers
don't match, as the example shows.

On the other hand, the order of Error and Maybe relative to each other isn't 
important.  Try it for yourself!

>>>>

\title{Error Reporting Parsers:  a Monad Transformer Approach}
\author{Matt Fenwick\email{mfenwick100@@gmail.com}}

@@

literal :: Eq t => t -> Parser t t
literal x = afilter (== x) item
\end{verbatim}
First, we define the Parser type, which says that a Parser is a function
of a list of tokens, producing a new list of tokens and a result
value, and possibly failing.
Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
consume a single token if available, and succeed otherwise.
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
For choosing between alternatives, the Alternative type class will do, so we create
an instance that is defined partly in terms of the Alternative instance of the
underlying result type, \verb+Maybe+.
Last are a couple of convenience combinators for creating specific token parsers.

Now for some examples:  first, let's see \verb+item+ in action.  It fails on empty lists,
and succeeds consuming one token otherwise:
\begin{verbatim}
*Main> getParser item ""
Nothing

*Main> getParser item "abcde"
Just ("bcde",'a')
\end{verbatim}

We can run parsers in sequence using the Applicative combinators.  First one parser
is run, then the second.  Note that the token position is handled such that the
second parser sees the token list output by the first parser, and that both parsers
must succeed in order for the entire parser to succeed:
\begin{verbatim}
*Main> getParser (item *> item) "abcde"
Just ("cde",'b')

@@

*Main> getParser (item *> item) "a"
Nothing
\end{verbatim}

Parsers built out of \verb+literal+ only succeed when the next token matches the given
one:
\begin{verbatim}
*Main> getParser (literal 'a') "abcde"
Just ("bcde",'a')

*Main> getParser (literal 'a') "bcde"
Nothing
\end{verbatim}

Using the Alternative combinators, we can create parsers that succeed if any of
their sub-parsers succeed.  This is what allows parsers to backtrack, trying
various alternatives if one fails:
\begin{verbatim}
*Main> getParser (literal 'a' <|> literal 'b') "abcde"
Just ("bcde",'a')

@@

First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer applied to Maybe, so the new
parser type is:
\begin{verbatim}
type Parser t a = StateT [t] Maybe a
\end{verbatim}

Now we can build the \verb+item+ parser using the combinators from MonadState, 
since its token lists are the 'state' that it modifies.
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}

@@

And a the \verb+literal+ combinator:
\begin{verbatim}
literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
literal c = satisfy ((==) c)
\end{verbatim}

@@

munch p = many0 (whitespace <+> comment) *> p
\end{verbatim}

Here are these three parsers in action:
\begin{verbatim}
*Main> runParser whitespace "\n \t   \tabc"
Just ("\n \t   \t","abc")

@@

The comment parser uses a new combinator, \verb+not1+, which is built out of a 
new type class, Switch, which models computations which can be switched from
successful to failing and vice versa:
\begin{verbatim}
class Switch f where
  switch :: f a -> f ()

@@

With \verb+munch+ in hand, we can now fix \verb+opencurly+ and try it out:
\begin{verbatim}
opencurly = munch $ literal '{'

*Main> map (runParser opencurly) [";abc\n  {def", "{def", "(def", "}def"]
[Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
\end{verbatim}

Excellent!  Now the \verb+opencurly+ parser correctly throws away whitespace 
and comments and succeeds, consuming one character, when the first non-junk 
character is an open curly brace, failing otherwise.
It's straightforward to define parsers for the other three brace tokens:
\begin{verbatim}
closecurly = munch $ literal '}'
openparen  = munch $ literal '('
closeparen = munch $ literal ')'
\end{verbatim}

@@

*Main> runParser symbol ";sdfasdfsa\n  123abc"
Nothing
\end{verbatim}

Note the use of \verb+many1+ to introduce repetition,
and we're again using \verb+munch+ to throw away leading junk.

\subsection{Syntactic structures}
Whereas our token parsers dealt with the syntactic primitives of the Woof grammar, 
the remaining parsers will have to implement grammar rules that combine the tokens
into syntactic structures.
The first combining form is a function application.  The rule says that it's delimited
by matching parentheses, in between which must appear one form as the operator, followed
by any number of forms as the arguments to which the operator is applied.  We can
say that like so:
\begin{verbatim}
application =
    openparen    *>
    pure AApp   <*>
    form        <*>
    many0 form  <*
    closeparen
\end{verbatim}

Note that, for now, we're doing all parser sequencing using the Applicative
combinators -- we don't need to use the monadic ones.  Also, we used \verb+pure+ 
to inject a value -- the function \verb+AApp+ -- into the parser.  \verb+pure+ 
has no effect on the tokens and always succeeds:
\begin{verbatim}
*Main> runParser (pure 3) ""
Just (3,"")

@@

Next, we move on to \verb+define+, which is also a straightforward translation 
from the grammar:
\begin{verbatim}
define =
    opencurly                    *>
    check (== "define") symbol   *>
    pure ADefine                <*>
    symbol                      <*>
    form                        <*
    closecurly
\end{verbatim}

Lambda is more complicated -- not only does it have additional syntax with its parameter list,
but we also need to ensure that the parameter names are unique.  We can do that using the 
\verb+check+ combinator again:
\begin{verbatim}
lambda = 
    opencurly                       *>
    check (== "lambda") symbol      *>
    opencurly                       *>
    pure ALambda                   <*>
    check distinct (many0 symbol)  <*>
    (closecurly                     *>
     many1 form                    <*
     closecurly)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
If the symbols are distinct, the parameter list subparser will succeed, and parsing will
continue; if there's a repeated symbol, the parsing will stop with a failure.

\subsection{Finishing touches}
Now all we have left to do is to define the \verb+special+, \verb+form+ (which 
we've already used to build the previous parsers), and \verb+woof+ parsers.  
The special parser is simple -- it parses either define or lambda:
\begin{verbatim}
special = define <+> lambda
\end{verbatim}

@@

except that the types don't match -- \verb+symbol+'s type parameter is a String, 
whereas the other two have ASTs, so we fix that by mapping the \verb+ASymbol+ 
function over \verb+symbol+:
\begin{verbatim}
form = fmap ASymbol symbol <+> application <+> special
\end{verbatim}

The final parser, \verb+woof+, needs to not only parse all the forms, but also 
make sure that the entire input is consumed.  The end of input check is done 
with the end parser:
\begin{verbatim}
end = switch item
\end{verbatim}

@@

*Main> runParser woof "a b c {define q (f x)},"
Nothing
\end{verbatim}
Note that the first two parses consume the entire input, while the last one 
fails because it can't recognize the trailing comma.

@@

runParser :: Parse e t a -> [t] -> Either e (Maybe (a, [t]))
runParser p xs = runMaybeT (runStateT p xs)
\end{verbatim}

A parser has three different possible results:  success, failure, and error:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))

@@

*Main> runParser (throwError "oops!") ""
Left "oops!"
\end{verbatim}
Although the \verb+<|>+ combinator can recover from failures, it can't recover from
errors.  Compare:
\begin{verbatim}
*Main> runParser (satisfy (== 'a') <|> satisfy (== 'b')) "babc"
Right (Just ('b',"abc"))

@@

\subsection{Type Class Preliminaries}
Unfortunately, the standard MonadError instance for the \verb+Either+ datatype does
not fit our needs -- it requires an \verb+Error+ constraint that we do not want.  To
solve this problem, we'll implement our own MonadError type class, based on the
standard one, and instances of it.  We'll need an instance for each member of
our stack -- Either, StateT, and MaybeT: 
\begin{verbatim}
class Monad m => MonadError e m | m -> e where
  throwError :: e -> m a
  catchError :: m a -> (e -> m a) -> m a

@@

\subsection{Token parsers}
By adding the position to the input, we've created a new problem 
for ourselves: our original token parsers operated on 
character lists, but we now need them to work on token lists.  
The cause of the problem is the \verb+literal+ combinator, so we
replace it with one for token lists:
\begin{verbatim}
character :: (MonadState [Token] m, Alternative m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}

And now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')

@@

ocurly = munch $ character '{'
ccurly = munch $ character '}'
oparen = munch $ character '('
cparen = munch $ character ')'
symbol = munch $ many1 char
  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
\end{verbatim}

The only differences were the replacement of \verb+literal+ with \verb+character+ 
and a couple uses of the \verb+chr+ accessor.  The \verb+munch+ combinator didn't 
have to change at all (although its inferred type did in fact change).

\subsection{Error messages}
Now we modify the remaining parsers to report accurate and useful errors.
Here are the error messages that we'll use to report each error:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eDefClose   =  "define: missing close curly"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eLamClose   =  "lambda: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

The MonadError type class provides two useful combinators, \verb+throwError+ for 
generating errors, and \verb+catchError+ for dealing with them.  We can ignore 
the second because we don't need to recover from errors, just report them, 
but when should an error be generated and what 
data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
introduce the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Alternative m) => e -> m a -> m a
commit err p = p <|> throwError err
\end{verbatim}

This combinator takes two arguments: 1. an error value, and 2. a monadic 
computation.  It first tries to run the computation, and if the computation 
fails, generates an error whose contents are the first argument.  The 
definition of 'failure' is provided by the Alternative instance of \verb+m+. So 
what data do we want to include in our errors?  A String describing the error 
and a token reporting the position should describe the error decently:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}

Now lets start with \verb+application+ as our first modified, error-reporting parser:
\begin{verbatim}
application =
    openparen                             >>= \open ->
    commit (eAppOper, open) form          >>= \op ->
    many0 form                            >>= \args ->
    commit (eAppClose, open) closeparen   >>
    return (AApp op args)
\end{verbatim}
This parser works by running the open-parenthesis parser, capturing its value 
in \verb+open+, and continuing with the parsing; if either \verb+form+ or 
close-parenthesis fails, an appropriate error can be generated which includes 
a string and the location of the open parenthesis.  However, if the open 
parenthesis is not successfully parsed, it's a failure but not an error.  
Examples:
\begin{verbatim}
*Main> runParser application (countLineCol "oops")
Right Nothing

@@

*Main> runParser application (countLineCol "(a b")
Left ("application: missing close parenthesis",('(',1,1))
\end{verbatim}

Although the previous version of the \verb+application+ parser was implemented 
with just applicative combinators, this version can't be:  \verb+open+ 
is used to build the following parsers, making this decidedly monadic.

@@

lambda =
    opencurly                                >>= \open ->
    check (== "lambda") symbol               >>
    commit (eLamParam, open) opencurly       >>= \p_open ->
    many0 symbol                             >>= \params ->
    (if distinct params 
        then return ()
        else throwError (eLamDupe, p_open))  >>
    commit (eLamPClose, p_open) closecurly   >>
    commit (eLamBody, open) (many1 form)     >>= \bodies ->
    commit (eLamClose, open) closecurly      >>
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}

Note that \verb+commit+ isn't used until we can be sure that we're in the right 
rule:  if, in the \verb+define+ parser, we committed to parsing the "define" 
keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
that failed.  Thus, if treating errors as unrecoverable, it's important not to 
place a \verb+commit+ where backtracking might be necessary to parse a valid 
alternative.  However, this isn't the case with how we've used it, since once 
we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule, and allowing backtracking would only destroy our ability to produce clean 
and accurate error reports.

\subsection{Error messages:  finishing touches}
To complete the error-reporting parser, we need to change a couple more details. 
First, we want an error reported when an open curly brace is found, but no 
special form can be reported.  To accomplish this, we can extend the \verb+special+ 
parser with a third alternative, which results in an error if it can parse
an open curly brace, but in a backtracking failure if not:
\begin{verbatim}
special = define <+> lambda <+> (ocurly >>= \o -> throwError (esName, o))
\end{verbatim}

The \verb+form+ parser doesn't have to be changed:
\begin{verbatim}
form = fmap ASymbol symbol <+> application <+> special
\end{verbatim}

When we can no longer parse any forms, instead of failing, we'd like to report an error if there's any input
left, and otherwise succeed.  To implement this parser, we'll first munch any leading whitespace or comments, 
as usual, and then we'll check the input:  if it's empty, great; if not, report an error at the location of the
first remaining token, along with the appropriate error message that we defined earlier:
\begin{verbatim}
endCheck = 
    many0 (whitespace <+> comment) *>
    get >>= \xs -> case xs of
                        (t:_) -> throwError (ewUnp, t)
                        []    -> pure ()
\end{verbatim}
Now our \verb+woof+ parser uses this \verb+endCheck+ parser after it has consumed all the forms that it can:
\begin{verbatim}
woof = many0 form <* endCheck
\end{verbatim}

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
*Main> runParser woof (countLineCol "{define fgh {lambda {f x} (f x)}}")
Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
                                     [AApp (ASymbol "f") [ASymbol "x"]])],
       []))
\end{verbatim}
Here's the proof showing that the parser correctly reports all of the errors 
covered by our spec:
\begin{verbatim}
*Main> let p = runParser woof . countLineCol

@@

runParser' :: Parser' t a -> [t] -> (Maybe a, [t])
runParser' p xs = runIdentity $ runStateT (runMaybeT p) xs
\end{verbatim}

We can compare it to the earlier parser:
\begin{verbatim}
*Main> runParser' (literal 'a') "aqrs"
(Just 'a',"qrs")

@@

*Main> runParser (literal 'a') "aqrs"
Just ('a',"qrs")
\end{verbatim}

The difference is that Maybe/State always returns a modified state, regardless
of the success of the computation, whereas State/Maybe only returns a modified
state if the computation is successful.  This means that backtracking doesn't
work right with Maybe/State -- tokens are consumed even when the parsers
don't match, as the example shows.

On the other hand, the order of Error and Maybe relative to each other isn't 
important.  But don't just take my word for it -- try it for yourself!
